---
title: "Medicare Physician Billing Patterns in RI and NH: A Comparative Analysis (2022–2023)"
author: "Shan Aziz"
date: "2025-06-31"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Introduction**

This report explores Medicare billing patterns among physicians practicing in Rhode Island (RI) and New Hampshire (NH) using the Medicare Physician & Other Practitioners by Provider dataset from the Centers for Medicare & Medicaid Services (CMS). We focus on providers with data available for both 2022 and 2023, allowing for a panel analysis at the provider-year level.

To ensure consistency and relevance, we restrict our analysis to physicians holding either a **Doctor of Medicine (MD)** or **Doctor of Osteopathic Medicine (DO)** degree, and whose practice location is within Rhode Island or New Hampshire. This allows for meaningful comparisons across states, specialties, and years.

**Key objectives:**

-   Summarize and compare **total submitted charges** and **total allowed charges** across the two states and years

-   Identify the **three most common physician specialties** and their distribution within each state

-   Estimate a **linear regression model** to test whether physicians in New Hampshire differ significantly in allowed charges compared to those in Rhode Island, controlling for specialty and year

-   Calculate the **correlation** between submitted charges in 2022 and allowed charges in 2023 to assess year-to-year billing consistency

The goal of this study is to uncover state-level and specialty-level patterns in Medicare billing and evaluate the stability of physician payment behavior across years.

```{r, message=FALSE, warning = FALSE}
# Loading libraries
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
library(tidyr)
library(patchwork)
library(scales)
library(stringr)
library(broom)
```

# Preliminary set up

##### Data source: [https://data.cms.gov/provider-summary-by-type-of-service/medicare-physician-other-practitioners/medicare-physician-other-practitioners-by-provider](https://data.cms.gov/provider-summary-by-type-of-service/medicare-physician-other-practitioners/medicare-physician-other-practitioners-by-provider){.uri style="color:blue;"}

```{r, message=FALSE}
# Loading both datasets
df_2022 <- read_csv("Medicare_Physician_Other_Practitioners_by_Provider_2022.csv")
df_2023 <- read_csv("Medicare_Physician_Other_Practitioners_by_Provider_2023.csv")

# Add year variable to each dataset before combining
df_2022$year <- 2022
df_2023$year <- 2023

# Combine both years into a single data frame
combined_df <- rbind(df_2022, df_2023)


```

### The unit of observation should be provider-year, with two observations per provider (2022 and 2023)

```{r}

# Get detailed breakdown of provider appearances
provider_counts <- table(combined_df$Rndrng_NPI)
provider_counts_df <- data.frame(
  Rndrng_NPI = names(provider_counts),
  count = as.numeric(provider_counts)
)

# See the distribution
table(provider_counts_df$count)

# Find providers in both years (balanced panel)
balanced_providers <- provider_counts_df$Rndrng_NPI[provider_counts_df$count == 2]
cat("Balanced panel providers:", length(balanced_providers), "\n")

# Find providers in only one year
single_year_providers <- provider_counts_df$Rndrng_NPI[provider_counts_df$count == 1]
cat("Single-year providers:", length(single_year_providers), "\n")

# Keep only providers appearing in both years
balanced_df <- combined_df[combined_df$Rndrng_NPI %in% balanced_providers, ]

# Verify: should have exactly 2 observations per provider
table(table(balanced_df$Rndrng_NPI))

```

### Filter for MD or DO and state is RI or NH

```{r}

# Advance cleaning step because of inconsistant MD and Do 

md_do_df <- balanced_df %>%
  filter(
    Rndrng_Prvdr_Crdntls %>%
      str_replace_all("M[,\\s]?D", "MD") %>%       # fix "M,D", "M D", etc.
      str_replace_all("D[,\\s]?O", "DO") %>%       # fix "D,O", "D O", etc.
      str_to_upper() %>%                           #uppercase
      str_replace_all("\\.", "") %>%               #Removes all periods
      str_replace_all("\\s+", " ") %>%             #Replaces multiple spaces with a single space
      str_trim() %>%                               #Removes leading and trailing spaces
      str_detect("\\bMD\\b")                       #\\b = word boundary in regex
  )

filtered_df <- md_do_df %>%
  filter(Rndrng_Prvdr_State_Abrvtn %in% c("RI", "NH"))

```

### **1. A short table of summary statistics comparing the Medicare billings of Rhode Island and New Hampshire physicians. Comparing the annual mean and standard deviation of total submitted charges and total allowed charges treated by physician state.**

```{r,message=FALSE}


# Summary Statistics
summary_stats <- filtered_df %>%
  group_by(State = Rndrng_Prvdr_State_Abrvtn, Year=year) %>%
  summarise(
    `Avg Submitted Charges` = mean(Tot_Sbmtd_Chrg, na.rm = TRUE),
    `SD Submitted Charges` = sd(Tot_Sbmtd_Chrg, na.rm = TRUE),
    `Avg Allowed Charges` = mean(Tot_Mdcr_Alowd_Amt, na.rm = TRUE),
    `SD Allowed Charges` = sd(Tot_Mdcr_Alowd_Amt, na.rm = TRUE),
  )

# Now I need to reshape this data to make a comparison chart
# Want to plot submitted vs allowed side by side
plot_data <- summary_stats %>%
  select(State, Year, `Avg Submitted Charges`, `Avg Allowed Charges`) %>%
  pivot_longer(cols = c(`Avg Submitted Charges`, `Avg Allowed Charges`),
               names_to = "Charge_Type", 
               values_to = "Amount")

# Reshape SDs
sd_data <- summary_stats %>%
  select(State, Year, `SD Submitted Charges`, `SD Allowed Charges`) %>%
  pivot_longer(cols = c(`SD Submitted Charges`, `SD Allowed Charges`),
               names_to = "Charge_Type", 
               values_to = "SD") %>%
  mutate(Charge_Type = gsub("SD", "Avg", Charge_Type))  # Match names to join

# Join average and SD data
plot_data <- left_join(plot_data, sd_data,
                       by = c("State", "Year", "Charge_Type"))

# Create a combined label for the x-axis to show state and year together
plot_data$State_Year <- paste(plot_data$State, plot_data$Year)

# Make the comparison chart
chart <- ggplot(plot_data, aes(x = State_Year, y = Amount, fill = Charge_Type)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Amount - SD, ymax = Amount + SD),
                position = position_dodge(0.9),
                width = 0.2, color = "black") +
  scale_fill_manual(values = c("Avg Submitted Charges" = "steelblue", 
                               "Avg Allowed Charges" = "green3")) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Healthcare Charges: Submitted vs Allowed",
       x = "State and Year",
       y = "Average Charges ($)",
       fill = "Charge Type") +
  theme_light()

# Show the chart
print(chart)

# Also show the detailed numbers in a table
kable(summary_stats, caption = "Summary Statistics by State and Year")


```

### Key Insights:

-   NH physicians submitted 50–70% more than RI physicians in both 2022 and 2023.

-   Despite higher submissions, allowed charges only showed a 2.4% decrease in NH and 0.44% increase in RI from 2022 to 2023

-   NH showed larger standard deviations, indicating more variation in physician billing behavior.

### **Three most common specialties in this data and reporting the proportion of doctors with these specializations within each state.**

```{r, message=FALSE}

# Find the top 3 most common medical specialties in overall data
top_specialties <- filtered_df %>%
  count(Rndrng_Prvdr_Type) %>%
  arrange(desc(n)) %>%
  head(3)

# Extract just the specialty names for filtering
top_3_names <- top_specialties$Rndrng_Prvdr_Type

# Filter my data to only include these top 3 specialties
data_top3 <- filtered_df %>%
  filter(Rndrng_Prvdr_Type %in% top_3_names)

# Now calculate what percentage each specialty makes up within each state
specialty_summary <- data_top3 %>%
  group_by(State = Rndrng_Prvdr_State_Abrvtn, Specialty = Rndrng_Prvdr_Type) %>%
  count() %>%
  group_by(State) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))

# Show the detailed breakdown in a table
kable(specialty_summary, caption = "Proportion of Top 3 Specialties by State")



# Now lets visualize this with pie charts for each state
# function to make consistent pie charts

make_pie_chart <- function(state_data, state_name) {
  ggplot(state_data, aes(x = "", y = Percentage, fill = Specialty)) +
    geom_col(width = 1) +
    coord_polar(theta = "y") +
    geom_text(aes(label = paste0(Specialty, "\n", Percentage, "%")), 
              position = position_stack(vjust = 0.7)) +
    labs(title = paste("Top 3 Specialties in", state_name)) +
    theme_void() +
    theme(legend.position = "none")
}

# Get data for each state separately
nh_data <- specialty_summary %>% filter(State == "NH")
ri_data <- specialty_summary %>% filter(State == "RI")

# pie charts for both states
nh_pie <- make_pie_chart(nh_data, "New Hampshire")
ri_pie <- make_pie_chart(ri_data, "Rhode Island")

# both charts side by side for easy comparison (special thanks to patchwork lib)
nh_pie + ri_pie
```

### **2. Regression that tests whether physicians who are in New Hampshire (vs. Rhode Island) have higher total allowed charges, controlling for provider type (i.e., specialty) and year.**

```{r}

# linear regression model
reg_model <- lm(Tot_Mdcr_Alowd_Amt ~ Rndrng_Prvdr_State_Abrvtn + Rndrng_Prvdr_Type + year, data = filtered_df)

# Tidy the model output
tidy_output <- tidy(reg_model)

# Accessing the 'term' column directly and use gsub for replacement
# Remove "Rndrng_Prvdr_Type"
tidy_output$term <- gsub("Rndrng_Prvdr_Type", "", tidy_output$term)

# Remove "Rndrng_Prvdr_State_Abrvtn"
tidy_output$term <- gsub("Rndrng_Prvdr_State_Abrvtn", "", tidy_output$term)

#output
kable(tidy_output,
      digits = 4,
      format = "latex",    # Explicitly set format for PDF output
      longtable = TRUE)    # multi-page tables

```

\

![[*https://itfeature.com/hypothesis/p-value-definition/*](https://itfeature.com/hypothesis/p-value-definition/)](values-of-p-Inference.jpg){width="345"}

**Interpretation of results**

We found a statistically significant difference in total allowed Medicare charges for physicians practicing in Rhode Island compared to the reference state. Specifically, Rhode Island providers had lower total allowed charges on average (Estimate = -19,083; p = 7.43e-08).

**Specialty Effect**

Several medical specialties were associated with significantly higher total allowed charges:

-   Medical Oncology (p = 0.000215)

-   Micrographic Dermatologic Surgery (p = 1.86e-07)

-   Dermatology (p = 0.001848)

-   Radiation Oncology (p = 0.005493)

-   Hematology-Oncology (p = 0.010535)

-   Interventional Pain Management (p = 0.022861)

-   Rheumatology (p = 0.019466)

Other specialties showed marginal significance (0.05 \< p \< 0.10), including:

-   Cardiac Surgery (p = 0.0501)

-   Hand Surgery (p = 0.0986)

-   Nephrology (p = 0.0837)

-   Orthopedic Surgery (p = 0.0869)

-   Pain Management (p = 0.0705)

These results suggest that specialty type is an important predictor of Medicare allowed charges.

**Year Effect**

There was no statistically significant linear trend in total allowed charges over time (p = 0.6401), indicating that Medicare payments did not meaningfully increase or decrease across the years analyzed.

**Model Fit**

The model explains approximately 19% of the variation in total allowed charges (Adjusted R² = 0.1865). This indicates that while the included variables contribute meaningfully, a large portion of the variation remains unexplained, suggesting the presence of other influential factors not captured in the model.

### **3. Specialty with the highest allowed charges (controlling for state and year)**

```{r}

# I want to find which medical specialty gets paid the most
# plan: extract coefficients -> filter for specialties -> find the highest one

# Get all coefficients from regression model
model_results <- coef(reg_model)

# I only care about specialty coefficients, not state/year effects
# These all have "Rndrng_Prvdr_Type" in their names
specialty_results <- model_results[grepl("Rndrng_Prvdr_Type", names(model_results))]

# This line looks at the names of the 'specialty_results'
# and replaces "Rndrng_Prvdr_Type" with an empty space ("").
names(specialty_results) <- gsub("Rndrng_Prvdr_Type", "", names(specialty_results))

# Sort them highest to lowest and grab the top one
highest_allowed_charges <- sort(specialty_results, decreasing = TRUE) %>% head(1)

# Printing highest allowed charges 
print(highest_allowed_charges)

```

Micrographic Dermatologic Surgery had the highest average total allowed charges, with an estimated value of \$542,258.59

### **4. Calculating the correlation between physicians’ total submitted charges in 2022 and total allowed charges in 2023.**

```{r}
# Create 2022 dataset with submitted charges
charges_2022 <- filtered_df %>%
  filter(year == 2022) %>%
  select(Rndrng_NPI, Tot_Sbmtd_Chrg) %>%
  rename(submitted_2022 = Tot_Sbmtd_Chrg)

# Create 2023 dataset with allowed charges
charges_2023 <- filtered_df %>%
  filter(year == 2023) %>%
  select(Rndrng_NPI, Tot_Mdcr_Alowd_Amt) %>%
  rename(allowed_2023 = Tot_Mdcr_Alowd_Amt)

# Only keep physicians who appear in both years for fair comparison
cor_df <- inner_join(charges_2022, charges_2023, by = "Rndrng_NPI")

# Calculate the correlation between 2022 submissions and 2023 allowances
correlation_result <- cor(cor_df$submitted_2022, cor_df$allowed_2023, use = "complete.obs")

# Show the result
print(paste("Correlation (r) =", round(correlation_result, 3)))
```

![[*https://financetrainingcourse.com/education/2011/04/correlation-correlation-coefficient-r/*](https://financetrainingcourse.com/education/2011/04/correlation-correlation-coefficient-r/){.uri}](download.jpeg){width="227"}

**Moderately strong positive correlation (r=0.659):** Physicians who submitted higher charges in 2022 tend to have higher allowed charges in 2023, suggesting consistent billing patterns across years.

### **Conclusion**

This analysis highlights important differences in Medicare physician billing patterns between Rhode Island and New Hampshire during 2022 and 2023. Physicians in New Hampshire consistently submitted higher charges than those in Rhode Island. However, the differences in allowed charges were relatively modest, suggesting that higher submissions do not necessarily result in proportionally higher reimbursements.

Physician specialty was a strong predictor of allowed charges. Micrographic Dermatologic Surgery had the highest average allowed charges when controlling for state and year. The regression results showed that Rhode Island physicians received significantly lower allowed charges on average compared to their New Hampshire counterparts.

No significant change in allowed charges was observed across the two years, indicating that Medicare payment levels remained stable during this period. Additionally, the moderately strong correlation between 2022 submitted charges and 2023 allowed charges suggests that physicians tend to maintain consistent billing patterns over time.

### **Final Remarks**

These findings offer insight into how physician location and specialty influence Medicare reimbursements. They can support efforts by policymakers, insurers, and healthcare administrators to better understand and address billing variation and promote equitable payment practices. Further analysis incorporating additional variables such as patient demographics, provider group size, and regional cost indices could help explain more of the variation in physician payments and inform future reforms.

------------------------------------------------------------------------

**Helpful Resources Links:**\
[*https://stackoverflow.com/questions/61635604/r-pivot-longer-and-ggplot-errorbar-with-two-name-key-columns*](https://stackoverflow.com/questions/61635604/r-pivot-longer-and-ggplot-errorbar-with-two-name-key-columns){.uri}

[*https://www.youtube.com/watch?v=Oe5O4LRj2rc&t=1972s*](https://www.youtube.com/watch?v=Oe5O4LRj2rc&t=1972s){.uri}*\
<https://stackoverflow.com/questions/61635604/r-pivot-longer-and-ggplot-errorbar-with-two-name-key-columns>*

[*https://www.geeksforgeeks.org/r-language/how-to-calculate-percentage-by-group-in-r/*](https://www.geeksforgeeks.org/r-language/how-to-calculate-percentage-by-group-in-r/){.uri}

[*https://r-graph-gallery.com/414-map-multiple-charts-in-ggiraph.html*](https://r-graph-gallery.com/414-map-multiple-charts-in-ggiraph.html){.uri}

[*https://www.sthda.com/english/articles/40-regression-analysis/163-regression-with-categorical-variables-dummy-coding-essentials-in-r/*](https://www.sthda.com/english/articles/40-regression-analysis/163-regression-with-categorical-variables-dummy-coding-essentials-in-r/){.uri}

[*https://stackoverflow.com/questions/30159162/linear-model-with-categorical-variables-in-r*](https://stackoverflow.com/questions/30159162/linear-model-with-categorical-variables-in-r){.uri}

[*https://www.statology.org/r-cor-function/*](https://www.statology.org/r-cor-function/){.uri}

[*https://www.geeksforgeeks.org/r-language/compute-the-correlation-coefficient-value-between-two-vectors-in-r-programming-cor-function/*](https://www.geeksforgeeks.org/r-language/compute-the-correlation-coefficient-value-between-two-vectors-in-r-programming-cor-function/){.uri}
